---
title: "Exploring the United State NOAA storm database"
output: html_document
---
##Introduction
The project aims to explore the storm database of the US National Oceanic and Atmospheric Administration (NOAA).These data include some characteristics in terms of major storms and atmospheric phenomena that occur in the United States.Also the damages and severe consequences that occurred in terms of deaths, injuries and property damage.So in this project, we will focus on the reasons that have led to these types mentioned above.

##Synopsis
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.This analysis present which types of events are most harmful with respect to population health and which have the greatest economic consequences.

##first Load required libraries
```{r,warning=FALSE,message=FALSE}
# set warning=FALSE and message=FALSE to stop seeing messages and warning
library(readr)
library(tidyr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

##loading Data
Download it from internet first
```{r,message=FALSE}
link <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url = link, destfile = "StormData")

```
Unzip Data and Read file as table format
```{r}
StormData <- read.csv(bzfile("StormData"),sep = ",",header=TRUE)
```

##Summary of Data
look at the data
```{r,results='markup'}
summary(StormData)
```

##Data Processing
At once we will prepare data for 2 questions
```{r}
#Select required variables only'effect of Events on health and economic' and drop the other.
StormDataSelected<-StormData[,c("BGN_DATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
#Subset Data '3columns ' that measures for Question 1
Health_harmful<-subset(StormDataSelected,!StormDataSelected$FATALITIES == 0 & !StormDataSelected$INJURIES == 
    0, select = c(EVTYPE, FATALITIES, INJURIES))
#Subset Data '4columns ' that measures for Question 2
Economic_harmful<-subset(StormDataSelected, !StormDataSelected$PROPDMG == 0 & !StormDataSelected$CROPDMG == 
    0, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

```

##1-Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
Data prpcessing 1.
Now we will providing the required data to answer the question regarding the effect of events on health.
```{r}
#for fatalities 
Health_harmful_death<-Health_harmful %>%
        group_by(EVTYPE) %>%
        summarise(FATALITIES_COUNT=sum(FATALITIES))%>%
        arrange(desc(FATALITIES_COUNT))
#select top 5 
Health_harmful_death<-Health_harmful_death[1:5,]
#for Injury
Health_harmful_injur<-Health_harmful %>%
        group_by(EVTYPE) %>%
        summarise(INJURIES_COUNT=sum(INJURIES))%>%
        arrange(desc(INJURIES_COUNT))
#select top 5 
Health_harmful_injur<-Health_harmful_injur[1:5,]
        
```

##Result for first question
plot the count of fatalities and injuries
```{r}
#use barplot from ggplot2 package
#plot top 5 events causing fatalities
Fatalities_plt<-ggplot(Health_harmful_death,
                       aes(x=EVTYPE,y=FATALITIES_COUNT,fill=EVTYPE))+
        geom_bar(stat = "identity",show.legend = FALSE)+
        xlab("Harmful Events")+ylab("num of Deaths")+
        ggtitle("Top 5 Events Causing Fatalities")+
        theme(axis.text.x = element_text(angle = 30))
#plot top 5 events causing Injuries
Injury_plt<-ggplot(Health_harmful_injur,
                   aes(x=EVTYPE,y=INJURIES_COUNT,fill=EVTYPE))+
        geom_bar(stat="identity",show.legend = FALSE)+
        xlab("Haemfull Events")+ylab("num of Injuries")+
        ggtitle("Top 5 Events Causing Injuries")+
        theme(axis.text.x = element_text(angle = 30))
#from "gridExtra" package we can divide two plots into 2col
grid.arrange(Fatalities_plt, Injury_plt, ncol = 2)

```

## 2-Across the United States, which types of events have the greatest economic consequences?
Data prpcessing 2.
Now we will providing the required data to answer the question regarding the effect of events on Economic.

```{r}
#to see count of each PROPDMGEXP and CROPDMGEXP levels 
table(Economic_harmful$PROPDMGEXP)
table(Economic_harmful$CROPDMGEXP)
```

```{r,results="hide"}
#we hidding the result of data is being too big
#select only levels required (K,k,M,m,B)
selected<-c("K","k","M","m","B")
Economic_harmful[Economic_harmful$PROPDMGEXP %in% selected,]
Economic_harmful[Economic_harmful$CROPDMGEXP %in% selected,]


#we convert Values to numeric 
#first replace pattern by numeric values by using gsub() function
Economic_harmful$PROPDMGEXP <- gsub("m", 1e+06, Economic_harmful$PROPDMGEXP, ignore.case = TRUE)
Economic_harmful$PROPDMGEXP <- gsub("k", 1000, Economic_harmful$PROPDMGEXP, ignore.case = TRUE)
Economic_harmful$PROPDMGEXP <- gsub("b", 1e+09, Economic_harmful$PROPDMGEXP, ignore.case = TRUE)
Economic_harmful$PROPDMGEXP <- as.numeric(Economic_harmful$PROPDMGEXP)
Economic_harmful$CROPDMGEXP <- gsub("m", 1e+06, Economic_harmful$CROPDMGEXP, ignore.case = TRUE)
Economic_harmful$CROPDMGEXP <- gsub("k", 1000, Economic_harmful$CROPDMGEXP, ignore.case = TRUE)
Economic_harmful$CROPDMGEXP <- gsub("b", 1e+09, Economic_harmful$CROPDMGEXP, ignore.case = TRUE)
#second convert them into numeric
Economic_harmful$CROPDMGEXP <- as.numeric(Economic_harmful$CROPDMGEXP)
Economic_harmful$PROPDMGEXP <- as.numeric(Economic_harmful$PROPDMGEXP)
#mutat column that contain the total damages
Economic_harmful$SumDam<-(Economic_harmful$PROPDMG * Economic_harmful$PROPDMGEXP)+
        (Economic_harmful$CROPDMG * Economic_harmful$CROPDMGEXP)
#third count total damges by each event
Economic_harmful<-Economic_harmful %>%
        group_by(EVTYPE) %>%
        summarise(TotalDMG=sum(SumDam))%>%
        arrange(desc(TotalDMG))
#select Top 5 only
Economic_harmful<-Economic_harmful[1:5,]

```

##Result for Second question
plot the count of Damages
```{r}
ggplot(Economic_harmful,
                           aes(x=EVTYPE,y=TotalDMG,fill=EVTYPE))+
        geom_bar(stat = "identity",show.legend = FALSE)+
        xlab("Harmfull Events")+
        ylab("sum of damages")+
        ggtitle("Top 5 Events Causing Economics Damages")
```

##Conclusion
After we make some analysis for this data to answer for both question ,we can found that the Tornado most harmfull affected to the population health , and the Flood have the greatest economic consequences.






